% !TEX spellcheck = en_US
% !TEX spellcheck = LaTeX
\documentclass[a4paper,10pt,english]{article}
\input{header}
\title{Lecture 07: Regenerative Processes and Renewal Theory}
\author{}
\begin{document}
\maketitle

\section{Regenerative Processes}
A stochastic process $Z = \{Z_t, t \geqslant 0 \}$ with the state space $E$ is said to be \textbf{regenerative} if there exists a sequence $S = \{S_n: n \in \N\}$ of stopping times such that 
\begin{enumerate}[(a)]
\item \textbf{regeneration times:} $S$ is a renewal process, 
\item \textbf{regenerative property:} for any $n,m \in \N$ and any bounded function $f : E^n \to \R$, we have
\begin{align*}
\E[f(Z_{S_m+t_1}, \dots, Z_{S_m+t_n})|Z_u, u \leqslant S_m] &= \E f(Z_{t_1},\dots,Z_{t_n}).
\end{align*}
\end{enumerate}
The definition says that probability law is independent of the past and shift invariant at renewal times. 
That is after each renewal instant, the process becomes a probabilistic replica of the process starting from zero. 
Let $F$ be the distribution of inter-renewal times, then for an open subset $A \subseteq E$ and $t \geqslant 0$, we define 
\begin{xalignat*}{3}
&K(t,A)= P\{S_1 > t, Z_t \in A\} && f(t) = P\{Z_t \in A\}.
\end{xalignat*} 
By the regeneration property applied at renewal instant $S_1$, we have
\begin{align*}
P\{Z_t \in A|S_1\} &= f(t-S_1) \text{ on } \{S_1 \leqslant t\}.
\end{align*}
Hence, we have the \textbf{renewal equation}
\begin{align*}
f(t) &= K(t,A) + \int_{0}^{t}dF(s)f(t-s) = K + F \ast f.
\end{align*}
We assume that the distribution function $F$ and the Kernel $K$ are known, and we wish to find $f$, and characterize its asymptotic behavior. 

\begin{thm} 
The renewal equation has a unique solution $f = m \ast g$, where $m(t) = \sum_{n \in \N_0}F_n(t)$ is the renewal function associated with the inter-renewal time distribution $F$. 
\end{thm}
\begin{proof}
It follows from the renewal equation that 
\begin{align*}
F\ast(m\ast g) &= \sum_{n \in \N}F_n \ast g = m \ast g - g.
\end{align*}
Hence, it is clear that $m \ast g$ is a solution to the renewal equation. 
For uniqueness, let $f$ be another solution, then $h = f - m \ast g$ satisfies $h = F \ast h$, and hence $h = F_n \ast h$ for all $n \in \N$. 
From finiteness of $m(t)$, it follows that $F_n(t) \to 0$ as $n$ grows. 
Hence, $\lim_{n\in\N}(F_n\ast h)(t) = 0$ for each $t$. 
\end{proof}

\begin{prop} 
Let $Z$ be a regenerative process with state space $E$. 
Then for any $t \geq 0$ and open $A \subseteq E$, 
\begin{align*}
P\{Z_t \in A\} &= \int_{0}^tdm(s)K(t-s,A),
\end{align*}
where $m$ is the renewal function and $K$ is the kernel function. 
\end{prop}

\begin{thm}[Key Lemma] Let $N(t)$ be a renewal process, with mean $m(t)$, \emph{iid} inter-renewal times $\{X_n\}$ with distribution function $F$, and $n^{\mathrm{th}}$ renewal instant $S_n$. Then,
\begin{equation*}
\Pr\{S_{N(t)}\leq s\}=\bar{F}(t)+\int_{0}^{s}\bar{F}(t-y)dm(y),\quad\quad t\geq s \geq 0.
\end{equation*}
\end{thm} 
\begin{proof} We can see that event of time of last renewal prior to $t$ being smaller than another time $s$ can be partitioned into disjoint events corresponding to number of renewals till time $t$. Each of these disjoint events is equivalent to occurrence of $n^{\mathrm{th}}$ renewal before time $s$ and $(n+1)^{\mathrm{st}}$ renewal past time $t$. That is,
\begin{equation*}
	\{S_{N(t)} \leq s\} = \bigcup_{n \in \N_0}\{ S_{N(t)} \leq s, N(t)=n\} = \bigcup_{n \in \N_0}\{ S_n \leq s, S_{n+1} > t\} .
\end{equation*}
Recognizing that $S_0 = 0$, $S_1 = X_1$, and that $S_{n+1} = S_n + X_{n+1}$, we can write
\begin{equation*}
	\Pr\{S_{N(t)} \leq s\} = \Pr\{X_1 > t\} + \sum_{n \in \N}\Pr\{ X_{n+1} + S_n > t, S_n \leq s\} .
\end{equation*}
We recall $F_n$, $n$-fold convolution of $F$, is the distribution function of $S_n$. Conditioning on $\{S_n = y\}$, we can write
\begin{align*}
	\Pr\{S_{N(t)} \leq s\} &= \bar{F}(t) + \sum_{n \in \N}\int_{y=0}^{s}\Pr\{ X_{n+1} > t - S_n, S_n \leq s| S_n = y\}dF_n(y),\\
	&= \bar{F}(t) + \sum_{n \in \N}\int_{y=0}^{s}\bar{F}(t-y)dF_n(y).
\end{align*}
Using monotone convergence theorem to interchange integral and summation, and noticing that $m(y) = \sum_{n\in\N}F_n(y)$, the result follows.
%\begin{flalign*}
%\Pr\{S_{N(t)} \leq s\} = \Pr\{X_1 > t\} +\sum_{n \in \N}\int_{y=0}^{s}\Pr\{ S_{n+1} > t|S_n=y\}dF_n(y)\\
%&= \Pr\{X_1 > t\} +\sum_{n \in \N}\int_{y=0}^{s}\Pr\{ X_{n+1} > t-y\}dF_n(y)\\
%&= \Pr\{X_1 > t\} +\sum_{n \in \N}\int_{y=0}^{s}\bar{F}(t-y)dF_n(y)\\
%&\stackrel{(a)}{=} \Pr\{X_1 >t) +\int_{y=0}^{s}\bar{F}(t-y)\sum_{n \in \N}dF_n(y),\\
%\end{flalign*}
%where the order of summation is exchanged by virtue of the fact that the summands are non-negative and the sum is finite (Fubini's theorem).
\end{proof}
\begin{shaded*}
%\begin{rem}\label{Remark:DensityLastRenewal} 
Key lemma tells us that distribution of $S_{N(t)}$ has probability mass at $0$ and density between $(0,t]$, that is,
\begin{align*}
\Pr\{S_{N(t)}=0\}&=\bar{F}(t),& dF_{S_{N(t)}}(y)&=\bar{F}(t-y)dm(y)~\quad 0 < y \leq t.
\end{align*}
%\end{rem}
%\begin{rem} 
Density of $S_{N(t)}$ has interpretation of renewal taking place in the infinitesimal neighborhood of $y$, and next inter-arrival after time $t-y$. To see this, we notice 
\begin{equation*}
dm(y)=\sum_{n \in \N}dF_n(y) =\sum_{n \in \N}\Pr\{n^{\text{th}} \text{renewal occurs in} (y,y+dy)\}.
\end{equation*}
Combining interpretation of density of inter-arrival time $dF(t)$, we get
\begin{equation*}
dF_{S_{N(t)}}(y)=\Pr\{\text{renewal occurs in }(y,y+dy) \text{ and next arrival after}~ t-y\}.
\end{equation*}
%\end{rem}
\end{shaded*}

\begin{lem} 
Let $F$ be the inter-renewal distribution such that $\inf\{x: F(x) = 1\} = \infty$, then for any $b > 0$ 
\begin{align*}
\sup_{t} \{m(t)-m(t-b) \}< \infty.
\end{align*} 
\end{lem}
\begin{proof}
We know that $m \ast F = m - 1$, so $m \ast (1- F) = 1$. 
Since the function $1-F$ is monotone, 
\begin{align*}
1 &= \int_0^tdm(s)[1-F(t-s)] \geq [m(t) - m(t-b)](1-F(b)),
\end{align*}
where $b$ is chosen so that $F(b) < 1$. Hence, the result follows. 
%\begin{align*}
%\sup_{t}[m(t) - m(t-b)] = \beta_b < \infty.
%\end{align*}
\end{proof}

%\begin{defn}[Lattice Random Variable] 
A non-negative random variable $X$ is said to be \textbf{lattice} if there exists $d \geq 0$ such that 
\begin{equation*}
\sum_{n\in\N}\Pr\{X = nd\} = 1.
\end{equation*}
For a lattice $X$, its period is defined as 
\begin{equation*}
d = \sup\{ d \in\R^+: \sum_{n\in\N}\Pr\{X = nd\} = 1 \}.
\end{equation*}
If $X$ is a lattice random variable, its distribution function $F$ is also called lattice.
%\end{defn}

%\subsection{Blackwell's Theorem}
\begin{thm}[Blackwell's Theorem] 
Let the inter-renewal times have distribution $F$, mean $\mu$, and the associated renewal function $m(t)$, such that $\inf\{x: F(x) = 1\} = \infty$. 
%Let $m(t)$ be the renewal function and the associated inter-arrival times with distribution $F$ and mean $\mu$. 
If $F$ is not lattice, then for all $a \geq 0$
\begin{equation*}
\lim_{t \to \infty} m(t+a) -m(t) = \frac{a}{\mu}.
\end{equation*}
If $F$ is lattice with period $d$, then 
\begin{equation*}
%\lim_{n \to \infty} \E[\text{number of renewals at } nd] = \frac{d}{\mu}.
\lim_{n \to \infty} m(t+d) - m(t) = \frac{d}{\mu}.
\end{equation*}
\end{thm}
\begin{proof} 
%From key renewal theorem, we have $\lim_{t \to \infty}m(t)/t = \frac{1}{\mu}$. For $\epsilon > 0$, pick $T$ such that for all $t > T$, we have
%\begin{align*}
%|m(t) - \frac{t}{\mu}| &\leq \epsilon/2,  |m(t+a) - \frac{t+a}{\mu}| &\leq \epsilon/2.
%\end{align*}
%Taking difference, we obtain for all $t \geq T$
%\begin{align*}
%|m(t+a) - m(t) - \frac{a}{\mu}| &\leq \epsilon.
%\end{align*}
We will not prove that 
\begin{equation}
\label{eq:LimitBlackwell}
g(a) = \lim_{t \to \infty} [m(t+a) - m(t)]
\end{equation}exists for non-lattice $F$. However, we show that if this limit does exist, it is equal to $a/\mu$ as a consequence of elementary renewal theorem. To this end, note that
\begin{equation*}
m(t+a+b) - m(t) = m(t+a+b) - m(t+a) + m(t+a) - m(t).
\end{equation*}
Taking limits on both sides of the above equation, we conclude that $g(a+b) = g(a) +  g(b)$. The only increasing solution of such a $g$ is
\begin{equation*}
g(a) = ca, \forall a > 0,
\end{equation*}
for some positive constant $c$. To show $c = \frac{1}{\mu}$, define a sequence $\{x_n, n \in \N\}$ in terms of $m(t)$ as 
\begin{equation*}
x_n = m(n) - m(n-1),~n \in \N.
\end{equation*}
Note that $\sum_{i=1}^nx_i = m(n)$ and $\lim_{n \in \N}x_n = g(1) = c$, hence we have
\begin{equation*}
%\lim_{n \to \infty}x_n = c \Rightarrow 
\lim_{n \in \N}\frac{\sum_{i=1}^nx_i}{n} = \lim_{n \in \N}\frac{m(n)}{n} \stackrel{(a)}{=} c,
\end{equation*}
where (a) follows from the fact that if a sequence $\{x_i\}$ converges to $c$, then the running average sequence $a_n = \frac{1}{n}\sum_{i=1}^n x_i$ also converges to $c$, as $n \to \infty$. 
Therefore, we can conclude $c = 1/\mu$ by elementary renewal theorem.

When $F$ is lattice with period $d$, the limit in~\eqref{eq:LimitBlackwell} doesn't exist. (See the following example). %\ref{eg:lattice limit}). 
However, the theorem is true for lattice trivially by elementary renewal theorem. Indeed, since $\frac{m(t)}{t} \to \frac{1}{\mu}$, we have that $m(nd)-m((n-1)d) \to \frac{1}{\mu}[nd-(n-1)d] = \frac{d}{\mu}.$
\end{proof}

\begin{shaded*}
%\label{eg:lattice limit}
For a trivial lattice example where the $\lim_{t\to \infty} m(t+a)-m(t)$ does not exist, consider a renewal process with $\Pr\{X_n = 1\} = 1$, that is, there is a renewal at every positive integer time instant with probability 1. Then $F$ is lattice with $d=1.$ Now, for $a=0.5$, and $t_n = n+(-1)^n 0.5$, we see that $\lim_{t_n \to \infty} m(t_n+a)-m(t_n)$ does not exist, and hence $\lim_{t \to \infty} m(t+a)-m(t)$ does not exist.
\end{shaded*}

%\begin{rem}
In the lattice case, if the inter arrivals are strictly positive, that is, there can be no more than one renewal at each $nd$, then we have that 
\begin{align}
\lim_{n\to \infty} P[{\text{renewal at nd}}] = \frac{d}{\mu}.
\end{align}
%\end{rem}

\end{document}