% !TEX spellcheck = en_US
% !TEX spellcheck = LaTeX
\documentclass[a4paper,10pt,english]{article}
\input{header}
\title{Lecture 28: Random walks}
\author{}

\begin{document}
\maketitle


\section{ GI/GI/1 Queueing Model}
Consider a $GI/GI/1$ queue. Customers arrive in accordance with a renewal process having an arbitrary interarrival distribution  $F$, and the service distribution $G$. %Let the interarrival times be $X_1,~X_2 \hdots $ and let the service times be $Y_1,~Y_2 \hdots$ and let $D_n$ denote the delay in queue of the $n^\text{th}$ arrival. 
\begin{prop}
\label{prop_gigi1}
Let $D_n$ be the delay in the queue of the $n^\text{th}$ customer in a GI/GI/1 queue with independent inter-arrival times $X_n$ and service times $Y_n$. 
Let $S_n$ be a random walk with \textit{iid} steps $U_n = Y_n - X_{n+1}$ for all $n \in \N$. 
%\begin{equation*}
%S_n =\sum_{i=1}^{n}Y_n-X_{n+1},~n \in \N.
%\end{equation*}
Then, we can write
\begin{align}
\label{GIGI1_prop_eqn}
\Pr\{D_{n+1} \geq c\} = \Pr\{S_j \geq c, \text{ for some }j \in [n]\}.
\end{align}
\end{prop}
\begin{proof}
The following recursion for $D_n$ is easy to verify
\begin{align*}
 D_{n+1} &= (D_n+Y_n-X_{n+1})1_{\{D_n+Y_n - X_{n+1} \geq 0\}} = \max\{0,D_n+U_n\}.
\end{align*}
%Using definition $U_n := Y_n-X_{n+1}$ for all $n \in \N$, we can re-write this as
%\begin{equation*}
%D_{n+1}=\max\{0,D_n+U_n\},~ n \in \N.
%\end{equation*}
Iterating the above relation with $D_1 = 0$ yields
\begin{align*}
D_{n+1}&= \max\{0,U_n+\max\{0,D_{n-1}+U_{n-1}\}\}=\max\{0,U_n,U_n+U_{n-1}+D_{n-1}\}.
\end{align*}
For the random walk $S_n$ with steps $U_n$, we can write delay in terms of random walk $S_n$ as
\begin{align*}
D_{n+1} &=\max\{0,S_n-S_{n-1},S_n-S_{n-2}, \ldots, S_n - S_0\}.
\end{align*}
Using the duality principle, we can rewrite the following equality for delay in distribution  
%where in the last step we have used the fact that $D_1=0$. Hence, for $c>0$,
\begin{align*}
D_{n+1} &=\max\{0,S_1,S_2, \ldots , S_n\}.
\end{align*}
%where the last equality follows from duality. Thus the following proposition holds.
\end{proof}
\begin{cor} If $\E U_n \geq 0$, then for all $c$, we have 
%\begin{align*}
$\Pr\{D_{\infty} \geq c\} \triangleq \lim_{n \in \N}\Pr\{D_{n} \geq c\} = 1$.
%\end{align*}
\end{cor}
\begin{proof}
It follows from Proposition~\ref{prop_gigi1} that $\Pr\{D_{n+1} \geq c\}$ is nondecreasing in $n$. 
Hence, by MCT the limit exists and is denoted by $\Pr\{D_{\infty} \geq c\}=\lim_{n \in \N}\Pr\{D_{n} \geq c\}$. 
Therefore, by continuity of probability, we have from~\eqref{GIGI1_prop_eqn}, that
\begin{align}
\label{random_walk_gigi1}
\Pr\{D_{\infty} \geq c\} &= \Pr\{S_n \geq c \text{ for some }n\}.
\end{align}
If $\E U_n = \E Y_n - \E X_{n+1}$ is positive, then by strong law of large numbers the random walk $S_n$ will converge to positive infinity with probability 1. 
The above will also be true when $E[U_n] = 0$, then the random walk is recurrent.
%Hence,
%\begin{equation*}
%Pr(D_{\infty} \geq c) =1,~ \forall c ~\text{if}~ E[Y]>E[X].
%\end{equation*}
\end{proof}
\begin{rem} Hence, we get that $\E Y_n< \E X_{n+1}$ implies the existence of a stationary distribution.
\end{rem}

\begin{prop}[Spitzer's Identity] Let $M_n=\max\{0,S_1,S_2, \ldots, S_n\}$ for $n \in \N$, then %We have the following proposition.
\begin{align*}
\E M_n &= \sum_{k=1}^{n} \frac{1}{k} \E S_k^+.
\end{align*}
\end{prop}
\begin{proof}
We can decompose $M_n$ as 
\begin{equation*}
M_n=1_{\{S_n>0\}}M_n+1_{\{S_n \leq 0\}}M_n.
\end{equation*}
We can rewrite first term in decomposition as, 
\begin{equation*}
1_{\{S_n>0\}}M_n= 1_{\{S_n>0\}}\max_{i \in [n]}S_i = 1_{\{S_n>0\}}(X_1+\max\{0,S_2 - S_1,\ldots, S_n - S_1\})
\end{equation*}
%By exchangeability, the joint distribution of $(X_1, \ldots, X_n)$ and $(X_n,X_1, \ldots, X_{n-1})$ are same.
Hence, taking expectation and using exchangeability, we get 
\begin{align*}
%\label{sptiz_1}
\E[M_n1_{\{S_n>0\}}] &= %\E 1_{\{S_n>0\}} X_1 + \E 1_{\{S_n>0\}} \max\{0,S_1,\ldots, S_{n-1}\} =   
\E[X_11_{\{S_n>0\}}]  + \E[M_{n-1}1_{\{S_n>0\}}].
\end{align*}
%By exchangeability, the joint distribution of $(X_1, \ldots, X_n)$ and $(X_n,X_1, \ldots, X_{n-1})$ are the same. 
%\begin{equation}
%\label{sptiz_2}
% E[ 1_{\{S_n>0\}} \max\{0,X_2,\hdots X_2+\hdots +X_n\}]=E[1_{\{S_n >0\}}M_{n-1}].
%\end{equation}
Since $X_i,S_n$ has the same joint distribution for all $i$,
\begin{align*}
\E S_n^+ &= \E[S_n1_{\{S_n>0\}}]=\E\sum_{i=1}^{n}X_i 1_{\{S_n>0\}}]=n\E[X_11_{\{S_n>0\}}].
\end{align*}
%Hence,
%\begin{equation}
%\label{sptiz_3}
%\E[X_11_{\{S_n>0\}}]=\frac{1}{n}\E[S_n 1_{\{S_n>0\}}] =\frac{1}{n}\E[S_n^+].
%\end{equation}
%From equations \ref{sptiz_1}, ~\ref{sptiz_2}, ~ \ref{sptiz_3}, we have that
Therefore, it follows that
\begin{align*}
\E[1_{\{S_n>0\}}M_n] &=\E[1_{\{S_n>0\}}M_{n-1}]+\frac{1}{n}\E[S_n^+].
\end{align*}
Also, $S_n \leq 0$ implies that $M_n=M_{n-1}$, it follows that
\begin{align*}
1_{\{S_n\leq 0\}}M_n=1_{\{S_n\leq 0\}}M_{n-1}.
\end{align*}
Thus, we obtain the following recursion,
\begin{align*}
\E[M_n]&=\E[M_{n-1}]+\frac{1}{n}\E[S_n^+].
\end{align*}
Result follow from the fact that $M_1 =S_1^+$.
%Upon recursion, we get
% \begin{equation*}
%\E[M_n]= \sum_{k=2}^{n} \frac{1}{k} \E[S_k^+]+\E[M_1].
%\end{equation*}
\end{proof}
\begin{rem} Since $D_{n+1} = M_n$, we have 
%From Proposition \ref{prop_gigi1}, with $M_n=\max\{0,S_1,\hdots S_n\}$
% \begin{equation*}
%Pr(D_{n+1} \geq c) = Pr(M_n \geq c).
%\end{equation*}
%Hence,
 %\begin{align*}
$\E[D_{n+1}] =\E[M_n] = \sum_{k=1}^{n} \frac{1}{k}E[S_k^+]$.
%\end{align*}
%From Spitzer's identity we see that
% \begin{equation*}
%E[D_{n+1}]= \sum_{k=1}^{n} \frac{1}{k}E[S_k^+].
%\end{equation*}
\end{rem}
\section{Martingales for Random Walks}
%Let
%\begin{align*}S_n = \sum_{k=1}^n X_k \quad n \geq 1\end{align*}
%denote a random walk. Then we have the following results.
\begin{prop}
A random walk $S_n$ with step size $X_n \in [-M,M]\cap \Z$ for some finite $M$ is a recurrent DTMC iff $\E X = 0$.
\end{prop}
\begin{proof}
If $\E X \neq 0$, the random walk is clearly transient since, it will diverge to $\pm \infty$ depending on the sign of $\E X$. 
Conversely, if $\E X =0$, then $S_n$ is a martingale. 
Assume that the process starts in state $i$. 
We define
\meq{3}{
&A = \{-M, -M+1,\cdots,-2,-1\}, &&A_j =\{j+1, \dots, j+M\}, ~j > i.
}
%\begin{align*}A_j = \{j, j+1,\cdots,j+M\}\end{align*}
Let $N$ denote the hitting time to $A$ or $A_j$ by random walk $S_n$. 
Since $N$ is a stopping time and $S_{N\wedge n} \leq |M|+j$, by optional stopping theorem, we have
\begin{align*}
\E_i[S_N] =\E_i[S_0] = i.
\end{align*}
Thus we have
\begin{align*}
i &= \E_i[S_N] %= \E_i[S_N|S_N \in A]\Pr\{S_N \in A\} + \E_i[S_N|S_N \in A_i]\Pr\{S_N \in A_j]\\
\geq -M\P_i\{S_N \in A\} + j(1-\P_i\{S_N \in A\}). 
\end{align*}
Rearranging this, we get a bound on probability of random walk $S_n$ hitting $A$ over $A_j$ as 
\begin{align*}
\P_i\{S_n \in A \text{ for some } n\} \geq \P_i\{S_N \in A\} \geq \frac{j-i}{j+M}.
\end{align*}
%Thus we have 
%\begin{align*}P[\mbox{process ever enters }A] \geq P[S_N \in A] \geq \frac{j-i}{j+M}\end{align*}
Taking limit $j \to \infty$, we see that for any $i \geq 0$, we have 
%\begin{align*}
$\P_i\{S_n \in A \text{ for some }n \} = 1$.
%\end{align*}
Similarly, taking $B = \{1,2,\cdots, M\}$, we can show that for any $i \geq 0$, 
%\begin{align*}
$\P_i\{S_n \in B \text{ for some }n \} = 1.$
%\end{align*}
Result follows from combining the above two arguments to see that for any $i \geq 0$,
\begin{align*}
\P_i\{S_n \in A\cup B \text{ for some } n\} = 1.
\end{align*}
\end{proof}
\begin{prop} 
Consider a random walk $S_n$ with mean step size $\E[X] \neq 0$. 
For $A,B > 0$, let $P_A$ denote the probability that the walk hits a value greater than $A$ before it hits a value less than $-B$. 
Then, for $\theta \neq 0$ such that $\E e^{\theta X_1} = 1$, we have
\begin{align*}
P_A \approx \frac{1-e^{-\theta B}}{e^{\theta A}-e^{-\theta B}}.
\end{align*}
Approximation is an equality when step size is unity and $A$ and $B$ are integer valued.
\end{prop}
\begin{proof}
For any $A,B > 0$, we can define stopping times
\meq{3}{
&T_A = \inf\{n \in \N: S_n \geq a\},&&T_{-B} = \inf\{n \in \N: S_n \leq -B\}.
}
We are interested in computing the probability 
\eq{
P_A &= \Pr\{T_A < T_{-B}\}.
}
%we wish to compute the probability $P_A$ that the walk hits at least $A$ before it hits a value $\leq -B$. %Let $\theta \neq 0$ s.t
%\begin{align*}
%\E[e^{\theta X}] = 1. 
%\end{align*}
Now let $Z_n = e^{\theta S_n}$. We can see that $Z_n$ is a martingale with mean 1. 
Define a stopping time $N = T_A \wedge T_{-B}$. 
%\begin{align*}
%N = \min \{S_n \geq A \mbox{ or } S_n \leq -B\}.
%\end{align*}
From Doob's Theorem, $\E[e^{S_N}] = 1$. 
Thus we get
\eq{
1 = \E[e^{\theta S_N}|S_N \geq A]P_A + \E[e^{\theta S_N}|S_N \leq -B](1-P_A).
}
We can obtain an approximation for $P_A$ by neglecting the overshoots past $A$ or $-B$. 
Thus we get
\meq{3}{
&\E[e^{\theta S_N}|S_N\geq A] \approx e^{\theta A},&&E[e^{\theta S_N}|S_N\leq -B] \approx e^{-\theta B}.
}
The result follows. 
\end{proof}

%\begin{cor} 
%Show that 
%\eq{
%\E N \approx \frac{AP_A - B(1-P_A)}{\E X}.
%}
%\end{cor}

\section{Application to G/G/1 Queues and Ruin}
\subsection{The G/G/1 Queue}
For the G/G/1 queue, we denote the inter-arrival time for $n$th arrival by $X_n$, and service time of $n$the customer by $Y_n$. 
Let $D_n$ be the delay in service for $n$th customer, and $S_n = \sum_{k=1}^{n}U_k$ be the location of random walk after $n$ steps with step size $U_n = Y_n - X_{n+1}$. 
From the equality in distribution of $D_{n+1}$ and $\max\{S_0, \dots, S_n\}$, we can obtain the limiting distribution of delay as 
\eq{
\Pr\{D_\infty \geq A\} = \Pr\{S_n \geq A \mbox{ for some } n\}.
}
%where 
%\begin{align*}S_n = \sum_{k=1}^n U_k, \quad U_k = Y_k-X_{k+1} \end{align*}
%Here $Y_i$ is the service time of the ith customer and $X_i$ is the interarrival duration between customer $i-1$ and customer $i$. 
Consider the case when $\E U = \E Y  - \E X < 0$, we choose $\theta > 0$ such that
\eq{
\E e^{\theta U}= \E e^{\theta(Y-X)} = 1.
}
From the previous equality and Gambler's ruin example, we get 
\eq{
\Pr\{D_\infty \geq A\} \leq e^{-\theta A}. 
}
\begin{prop} 
Consider a GI/M/1 queue with independent and \textit{iid} random inter-arrival times $\{X_n: n \in \N\}$ and exponential service times $\{Y_n: n \in \N\}$ with rate $\mu$.  
We choose $\theta$ such that $\E e^{\theta (Y_n - X_{n+1})} = 1$, then the limiting delay distribution for the GI/G/1 queue is given by 
\eq{
\Pr\{D_\infty \geq A\} = \frac{\mu - \theta}{\mu}e^{-\theta A}.
}
\end{prop}
\begin{proof} 
From the choice of $\theta$, it's clear that $\E e^{\theta S_n}$ is a martingale. 
For a fixed $A, B > 0$, we define stopping times $T_A, T_{-B}, N = T_{A}\wedge T_{-B}$. 
Since $|S|_{N} \leq \max{A,B}+ \sup_{n}|X_n|$, using optional stopping theorem, we can write 
\eq{
1 &= \E[e^{\theta S_N}|S_N\geq A]\Pr\{T_A < T_{-B}] + \E[e^{\theta S_N}|S_N\leq -B]\Pr\{T_A > T_{-B}\}.
}
We notice that conditioned on $N = n$ and $X_{n+1}- S_{n-1}$, we have 
\eq{
S_N = Y_n - X_{n+1} + S_{n-1}. 
} 
We focus on first term on right hand side, to get 
\eq{
\E[e^{\theta S_N}|S_N\geq A] &= \E[\E[e^{\theta S_N}| N = n, X_{n+1}- S_{n-1}]|S_N\geq A] \\
&=  \E[e^{\theta (X_{n+1} - S_{n-1}) }\E[e^{\theta Y_n}| N = n, X_{n+1}- S_{n-1}]|S_N\geq A]\\
&= \E[e^{\theta (X_{n+1} - S_{n-1}) }\E[e^{\theta Y_n}|Y_n\geq (X_{n+1}-S_{n-1})+ A]| N = n, X_{n+1}- S_{n-1}].
}

\end{proof} 

%Now the exact distribution of $D_\infty$ can be calculated when services are exponential. 
%Hence assume $Y_i \sim exp(\mu)$. 

Let us compute $E[e^{\theta S_N}|S_N \geq A]$ first. Let us condition this on $N=n$ and $X_{n+1} - \sum_{i=1}^{n-1} (Y_i - X_{i+1}) = c$. By the memoryless property, the conditional distribution of $Y_n$ given $Y_n > c+A$ is just $c+A$ plus an exponential with rate $\mu$. Thus we get
\begin{flalign*}
E[e^{\theta S_N}|S_N \geq A] &= E[e^{\theta(A+Y)}] \\
&=\frac{\mu e^{\theta A}}{\mu - \theta}
\end{flalign*}
Now substituting back, we get
\begin{flalign*}
1 &= \frac{\mu e^{\theta A}}{\mu - \theta}P[S_n\mbox{ crossed $A$ before $-B$}] \\
&+ E[e^{\theta S_N}|S_N\leq -B]P[S_n\mbox{ crossed $-B$ before $A$}]
\end{flalign*}
Now as $\theta > 0$, let $B\to \infty$ to get
\begin{align*}1 = \frac{\mu e^{\theta A}}{\mu - \theta} P[S_n \mbox{ ever crosses }A]\end{align*}
And hence

\subsection{A Ruin Problem}
Suppose claims made to an insurance company follow a renewal process with iid interarrival times $\{X_i\}$. Let the values of the claims also be iid and independent of the renewal process $N(t)$ of their occurence. Let $Y_i$ be the ith claim value. Thus the total value of claims till time $t$ is $\sum_{k=1}^{N(t)}Y_i$. Now let us suppose the insurance company receives money at constant rate $c$ per unit time, $c>0$. We wish to compute the probability of the insurance company, starting with capital $A$, will eventually be wiped out or \textbf{ruined}. Thus we require
\begin{align*}p = P\left\{ \sum_{k=1}^{N(t)}Y_i > ct + A \mbox{ for some } t\geq 0\right\}\end{align*}
As an assignment, show that the company will be ruined if $E[Y] \geq cE[X]$. So let us assume that $E[Y] < cE[X]$. Also the ruin occurs when a claim is made. After the $n$th claim, the company's fortune is
\begin{align*}A + c\sum_{k=1}^n X_k -\sum_{k=1}^n Y_k \end{align*}
Letting $S_n = \sum_{k=1}^n Y_i - cX_i$ and $p(A) = P[S_n > A \mbox{ for some }n]$. As $S_n$ is a random walk, we see that 
\begin{align*}p(A) = P[D_\infty > A]\end{align*}

Now the results from the G/G/1 queue apply.

\section{Blackwell Theorem on the Line}
Let $S_n$ denote a random walk where $0<\mu=E[X] < \infty$. Let 
\begin{align*}U(t)=\#\{n: S_n \leq t\} = \sum_{n=1}^\infty I_n\end{align*}
Where $I_n=1$ if $S_n \leq t$ and zero else. Observe that if $X_n$ are nonnegative, then $U(t) = N(t)$. Let $u(t) = E[U(t)]$. Now we prove an analog of Blackwell Renewal Theorem. 

\begin{thm}
\textbf{(Blackwell renewal theorem)} If $\mu > 0$ and $X_i$ are not lattice, then
\begin{align*}u(t+a) - u(t) \to a/\mu \quad t\to \infty \quad \mbox{for }a>0\end{align*}
\end{thm}
Let us define a few concepts. We say an \textbf{ascending ladder variable of ladder height $S_n$} occurs at time $n$ when
\begin{align*}S_n > \max(S_0,S_1,\cdots, S_{n-1})\end{align*}
where $S_0 = 0$. We may deduce that since $X_i$ are iid random variables, then the random variables $(N_i,S_{N_i}-S_{N_{i-1}})$ are iid; where $N_i$ denotes the time between the $(i-1)$th and ith random variable. We may analogously define descending ladder variables. Now let $p(p_*)$ denote the probability of ever achieving an ascending/descending ladder variable.
\begin{align*}p = P\{S_n > 0 \mbox{ for some }n\},\quad p_* = P\{S_n < 0 \mbox{ for some }n\}\end{align*}
At each ascension/descension there is a probability $p$ (resp $p_*$) of achieving another one. Hence the number of ascensions/descensions is geometrically distributed. The number of ascending ladder variables (ascensions) will have finite mean iff $p < 1$. Now as $E[X] > 0$, by SLLN, we deduce that $w.p.1$, there will be infinitely many ascending ladder variables but finitely many descending ones. That is $p =1$ and $p_* < 1$.

\begin{proof}
The successive ascending ladder heights are a renewal process. Let $Y(t)$ be the excess time. Now given the value of $Y(t)$, the distribution of $U(t+a) - U(t)$ is independent of $t$. (Why?). Hence let us denote
\begin{align*}E[U(t+a) - U(t)|Y(t)] = g(Y(t))\end{align*}
for some function $g$. Now taking expectations yields
\begin{align*}u(t+a) - u(t) = E[g(Y(t))]\end{align*}

Now since $Y(t) \to^d Y_\infty$ where $Y_\infty$ has the equilibrium distribution, we have $E[g(Y(t))] \to E[g(Y_\infty)]$. The result would be true if we show $g$ is continuous and bounded. We leave that as an exercise. For now, we deduce that the limit exists. Let
\begin{align*}h(a) = \lim_{t \to \infty}u(t+a) - u(t)\end{align*}
This also implies $h(a+b) = h(a) + h(b)$. Thus for some constant $c$, 
\begin{align*}h(a) = ca\end{align*}

Now to get $c$, let $N_t$ denote the first $n$ for which $S_n > t$. If $X_i$ are upper bounded by $M$, then
\begin{align*}t < \sum_{i=1}^{N_t} X_i \leq t+M\end{align*}
Taking expectations, and using Wald's Lemma, yields
\begin{align*}t < E[N_t]\mu \leq t+M\end{align*}
Thus 
\begin{align*}\frac{E[N_t]}{t} \to \frac{1}{\mu}\end{align*}
If $X_i$ are unbounded, use the truncation arguments done while proving Elementary renewal theorem. Now $U(t)$ can be expressed as
\begin{align*}U(t) = N_t -1 +N_t^*\end{align*}
where $N_t^*$ is the number of times $S_n \leq t$ after having crossed $t$. Since $N_t^*$ is not greater than the number of points occuring after $N_t$ when the random walk is less than $S_{N_t}$, we get
\begin{align*}E[N_t^*] \leq E[\mbox{number of $n$ such that $S_n < 0$}]\end{align*}

Hence if we argue that RHS of above is finite, then
\begin{align*}\frac{u(t)}{t} \to \frac{1}{\mu}\end{align*}

From the first proposition in Random walks, we have $E[N] <\infty$ where $N$ is the first value of $n$ for which $S_n > 0$. At time $N$, with positive probability $1-p^*$, no future value of random walk will fall below $S_N$. Thus,
\begin{align*}E[\mbox{number of $n$ where $S_n < 0$}] \leq \frac{E[N|X_1<0]}{1-p^*} < \infty\end{align*}

Now follow the steps illustrated in the Blackwell renewal theorem (original) proof to arrive at the desired result.

\end{proof}
\end{document}
