\documentclass[a4paper,10pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[colorlinks,urlcolor=blue]{hyperref}

\title{SPQT- Asst 3 (DTMC)}
\author{Prof. Parimal Parag}

\begin{document}
\maketitle
\section{Notations}
DTMC means Discrete Time Markov Chain. TPM means Transition Probability Matrix. The entries of the TPM are given as $P_{ij}$. 
\begin{enumerate}
	\item There are a total of $N$ balls in urns A and B. At step $k$, one of the $N$ balls is chosen at random i.e. w.p. $1/N$. Then one of the urns A or B is picked (prob. of picking A is $p$) and the chosen ball is put in that urn. Let $X_k$ denote the number of balls in urn A at step $k$. Then
	\begin{enumerate}
		\item Show that $\{X_k\}$ is a DTMC.
		\item Determine the state space and TPM.
		\item Find if it is irreducible or not.
		\item Compute $\lim_{n\to \infty} P_{ij}^n$ for all $i,j$.
	\end{enumerate}
	
	\item Show that for a finite state aperiodic irreducible DTMC, $P^n_{ij} > 0$ $\forall i,j$ for all n large enough.
	
	\item Let $P$ be the TPM of an irreducible DTMC such that $P^2=P$. Show that the DTMC is aperiodic and $P_{ij} = P_{jj}$ for all $i,j$.
	
	\item Let a DTMC have $r < \infty$ states. Show that
	\begin{enumerate}
		\item If state $k$ can be reached from state  $j$, then it can be reached in at most $r-1$ steps.
		\item If $j$ is recurrent state, then $\exists \alpha, \quad 0 < \alpha < 1$ s.t. for $n>r$ the probability that first return to state $j$ (from $j$). occurs after n transitions is $\leq \alpha^n$.
	\end{enumerate}
	
	\item Let $P$ be a TPM that is doubly stochastic (Both Rowsum and Columnsums are 1). Show that if the underlying DTMC is finite and irreducible, then it's stationary probability $\pi$ satisfies $\pi(i) = \pi(j)$ for all $i,j$.
		
	\item Let a DTMC have state space $\{0,1,2,3,4,5\}$ with TPM
	\[
	 \begin{array}{c| c c c c c c}
	P_{ij}&0&1&2&3&4&5 \\ \hline
	0&1/3 & 2/3 & 0& 0&0&0 \\
	1&2/3&1/3&0&0&0&0 \\
	2&0&0&1/4&3/4&0&0\\
	3&0&0&1/5&4/5&0&0\\
	4&1/4&0 &1/4 &0 & 1/4 & 1/4 \\
	5&1/6 &1/6 &1/6 &1/6 &1/6 &1/6 	
	\end{array} \]
		Find all closed sets, show transient states and calculate $\lim_{n \to \infty}P_{5i}^n$ for all $i$.
		
		\item Consider a recurrent DTMC starting in state $0$. Let $m_i$ denote the expected no. of time periods it spends in state $i$ before returning to $0$. Use Wald's equation to show that \[ m_j = \sum_i m_i P_{ij}, \quad j >0,\quad  m_0 = 1\]
		
		\item Let $X_i$ be independent random variables such that $P[X_i = j] = \alpha_j, j \geq 0$. Say a record occurs at time n if $X_n > \max(X_1,X_2, \cdots, X_{n-1})$ where $X_0 = -\infty$. If a record occurs at time $n$, $X_n$ is a record value. Let $R_i$ denote the ith record value.
		\begin{enumerate}
			\item Argue that $\{ R_i, i \geq 1\}$ is a DTMC. Compute its TPM. 
			\item Let $T_i$ denote the time between the ith and (i+1)th record. Is $\{T_i\}$ a DTMC? What about $\{(R_i,T_i)\}$? Compute TPM wherever applicable.
			\item Let $S_n = \sum_{i=1}^n T_i, \quad n\geq 1$. Argue that $\{S_n, n \geq 1\}$ is a DTMC. Find it's TPM.
		\end{enumerate}
		
		\item Let $P$ be the TPM of a DTMC with state space $\{0,1,2,\cdots, N\}$. Let $\{0,N\}$ be absorbing states and $\{1,\cdots, N-1\}$ be irreducible transient states. Form another TPM $\overline{P}$ with $\overline{P}(0,1) = 1 = \overline{P}(N,1)$ and other elements as in $P$. Then $\overline{P}$ is TPM of an irreducible DTMC with stationary distribution $\pi$. Show that if original DTMC (w/ TPM $P$) starts in state 1, it's mean time to absorption in $\{0,N\}$ is \[\frac{1}{\pi(0) +\pi(N)} -1\]
\end{enumerate}

Mail any typos to \url{gautamshenoydmc@gmail.com}.
\end{document}