\documentclass[a4paper,10pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[hidelinks]{hyperref}

\title{SPQT- Asst 1 (Poisson Processes)}
\author{Prof. Vinod Sharma}

\begin{document}
\maketitle
\begin{enumerate}
	\item An item has a random lifetime with exponential distribution with parameter $\lambda$. When it fails, it is replaced immediately by another identical item and so on. Let $N_t$ denote the number of failures till time $t$. Prove that $N_t$ is a Poisson Process. Find the mean and variance of the time till the fifth item fails.
	\item Let $A_1,A_2, \cdots, A_n$ be disjoint intervals of positive line with union $B$. Let $a_1,a_2, \cdots, a_n$ be their respective lengths. Let $b=\sum_{i=1}^n a_i$. Then for $k_1 + k_2+\cdots + k_n = k$, show that
\[ P\left[ N_{A_1} = k_1, N_{A_2} = k_2, \cdots, N_{A_n} = k_n | N_B = k\right] = \frac{k!}{k_1! k_2!\cdots k_n!} \prod_{i=1}^n\left(\frac{a_i}{b}\right)^{k_i}\]

\item A department store has 3 doors. Arrivals at each door form a Poisson Process with rates $\lambda_1 = 110$, $\lambda_2 = 90$ and $\lambda_3 = 160$ customers per hour. $30\%$ of all customers are male. Prob. that a male purchases something is $0.8$ whereas for a female, it is $0.1$. An average purchase is worth Rs. $4.50$. Assume all random variables are independent. What is the average worth of total sales in 10 hrs? What is the probability that the third female to purchase anything comes in the first 15 minutes? Also compute the expected time of her arrival.

\item For a Poisson Process $N_t$ with rate $\lambda$, show that ($S_i$ is the arrival instants) 
\[ P\left[S_1=t_1, \cdots, S_n=t_n | N_t = n\right] = \frac{n!}{t^n}\]
where $0 < t_1 < t_2 < \cdots t_n < t$.

\item Customers arrive according to a Poisson Process with parameter $\lambda$. There is a waiting cost of c per customer per unit time. The customers wait at the facility until they are dispatched. Dispatching occurs at epochs $T, 2T, \cdots$ for a fixed positive constant T. Any customer present at time $kT$ is dispatched at $kT$ instantaneously. There is a dispatching cost of $\beta$ per customer. 
\begin{enumerate}
	\item What is the total dispatch cost till time T?
	\item What is the mean total customer wait till time T?
	\item What value of T minimizes the mean total customer cost and dispatch cost per unit time?
	\end{enumerate}

\item Let $N(t)$ be a Poisson Process with parameter $\lambda$. Let $S_i$ be the arrival epochs. Compute $E[S_5|N(t)=3]$.

\item Let $N_1(t), N_2(t)$ be two independent Poisson processes with parameters $\lambda_1$ and $\lambda_2$ respectively. Let $S_n^{(1)}$ and $S_n^{(2)}$ be the respective arrival epochs. Compute
\begin{enumerate}
	\item $P[S_1^{(1)} < S_1^{(2)}]$
	\item $P[S_2^{(1)} < S_2^{(2)}]$
\end{enumerate}

\item Shocks occur to a system according to a Poisson Process of intensity $\lambda$. Each shock causes damage and these damages accumulate. Let $N_t$ be the number of shocks till time $t$ and $Y_i$ be the damage caused by the $i^{th}$ shock. Assume that $Y_i$ is independent of $N(t)$ and that it is distributed geometrically as $P[Y_i = n] = \alpha(1-\alpha)^{n-1}, \quad n=1,2,\cdots$. Define $X(t)$, the total damage till time $t$ as
\[X(t) = \sum_{i=1}^{N(t)} Y_i\]
Assume that system functions until $X(t) > a$ where $a>0$. Compute the mean time till failure. 
\item (M/M/1 Queue) To a single server queue, arrivals $(N_t)$ come as per a Poisson Process with rate $\lambda$. Customers require an iid service of exponentially distributed time of rate $\mu$ after which, they leave the queue. Let $q_t$ be the queue length at time $t$ and $D_t$ the number of departures till time $t$. Hence $q_t=N_t - D_t$.
\begin{enumerate}
	\item Calculate $P[D_t=m|N(t)=n]$
	\item Calculate $P[q_t=m|N(t)=n]$
	\item Calculate $P[q_t=m], E[q_t], E[q_t^2]$
	\item Let $Q_n$ be the queue length when the nth customer arrives (excluding itself). Basically, it is the number of customers in queue an arriving customer sees. Calculate $P[Q_n = n| Q_{n-1} = m]$
\end{enumerate}
\end{enumerate}

Mail any typos to \url{gautamshenoydmc@gmail.com}.
\end{document}