\documentclass[a4paper,10pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[colorlinks,urlcolor=blue]{hyperref}

\title{SPQT- Asst 6 (Random Walks)}
\author{Prof. Parimal Parag}

\begin{document}
\maketitle
\section{Notes}
\begin{enumerate}
	\item A weak ascending ladder epoch for a RW is $\inf \{n \geq 1: S_n \geq 0\}$ and the strong (or strict) ascending ladder epoch is $\inf \{n \geq 1: S_n > 0\}$.
	\item $X=^d Y$ means $X$ and $Y$ have the same distribution.
\end{enumerate}
\section{Questions}
\begin{enumerate}
	\item Let $\{X_i\}$ be iid  with $E[X_1]> 0$. Using Elementary renewal theorem on ascending ladder heights, show that
	\[\frac{E[\tau (u)]}{u} \sim \frac{u}{E[X_1]}\]
	where $\tau (u) = \inf \{n>0: S_n > u\}$.
	\item For a random walk $S_n$, let $\tau^w_+$ be its weak ascending ladder epoch and $\tau_+$ be the strong ascending ladder epoch. Let $\rho = P[S_{\tau^w_+} = 0] < 1$. Show that 
	\[\tau_+ = \sum_{k=0}^{N-1}T_k\]
	where $T_0, T_1, \cdots $are independent and 
	\begin{flalign*}
	T_0 &=^d \tau_+^w \quad \mbox{if } S_{\tau^w_+} > 0 \\
	T_1, T_2, \cdots &=^d \tau_+^w \quad \mbox{if } S_{\tau^w_+} = 0
	\end{flalign*}
	Here $N$ is independent of $T_i$ for every $i$ and 
	\[P[N=n] = \rho^{n-1}(1-\rho)\]
	
	\item For a random walk $S_n$, let $M=\sup_{k \geq 0}S_k$. 
	\begin{enumerate}
		\item Consider the Wiener Hopf equation given by
		\[Z(x) = z(x) + \int_0^x Z(x-y)d\mu(y)\] 
		where $\mu$ is a probability measure. Show that $P[M > x]$ satisfies this.
		\item If $E[X_1]=0, Var(X_1)=1$, we can show that $M_n/\sqrt{n} \to^d \mathcal{N}(0,1)$. From this, show that for a GI/GI/1 queue,
		\[W_n/\sqrt{n} \to^d \mathcal{N}(0,1) \]
		if $\rho = 1$ and second moment of interarrival time and service time are finite.
		\item Show that for $\rho > 1$, $W_n/n \to^d (\rho - 1)/E[A]$ where $E[A]$ is the mean interarrival time.
	\end{enumerate}
	
	\item Let $S_n = \sum_1^n X_i$ be a random walk. Let $\lambda_i,\quad i>0$ denote the probability that a ladder height is $i$. 
	\begin{enumerate}
		\item Show that if
		\[P[X_i = j] = \left\{\begin{array}{lc}
			q,  & j=-1 \\
			\alpha_j, & j \geq 1
		\end{array}\right.\]
		where $q=1- \sum_1^\infty \alpha_j$, then
		\[\lambda_i = \alpha_i + q(\lambda_{i+1} +\lambda_1\lambda_i)\]
		\item If $P[X_i=j] = 1/5$ for $j=-2,-1,0,1,2$, show that 
		\[\lambda_1 = \frac{1+\sqrt{5}}{3+\sqrt{5}}, \quad \lambda_2 = \frac{2}{3+\sqrt{5}}\]
	\end{enumerate}
	
	\item Let $X_1,\cdots, X_n$ be equally likely to be any of the $n!$ permutations of $(1,2,\cdots, n)$.Show that
	\[P\left[\sum_{j=1}^n jX_j \leq a\right] = P\left[\sum_{j=1}^n jX_j \geq n(n+1)^2/2-a\right]\]
\end{enumerate}

Mail any typos to \url{gautamshenoydmc@gmail.com}.
\end{document}