\documentclass[a4paper,10pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[hidelinks]{hyperref}

\title{SPQT- Asst 2 (Renewal Theory)}
\author{Prof. Parimal Parag}

\begin{document}
\maketitle
\section{Notation}
Let $\{X_i\}$ be iid non-negative random variables with distribution $F$. Let $S_n = \sum_{k=1}^n X_k$. Let $N(t)$ be the no. of arrivals till time $t$, $m(t) = E[N(t)]$ and $\mu = E[X_1]$. Also the presence of a ``bar'' on a cdf term indicates complimentary cdf. e.g. $\overline{F}(u) = 1 - F(u)$.

\section{Questions}
\begin{enumerate}
	\item Show that $P[X_{N(t) + 1} > x] \geq \overline{F}(x)$. Hence deduce that $E[X_{N(t)+1}^m] \geq E[X_1^m]$, where $m$ is any positive integer. Compute $P[X_{N(t) + 1} > x]$ when $X_i$ are iid exponential($\lambda$).
	\item Prove the renewal equation
	\[m(t) = F(t) + \int_{0}^t m(t-u)dF(u)\]
	If $F$ is uniform on $[0,1]$, show that $m(t) = e^t -1 \quad 0\leq t \leq 1$.
\item Consider a single server bank in which customers arrive as a Poisson process with rate $\lambda$. However an arrival only enters the bank if the server is free when the customer arrives. Let $G$ denote the service distribution.
\begin{enumerate}
	\item What fraction of time is the server busy?
	\item At what rate do customers enter the bank?
	\item What fraction of potential customers enter the bank?
\end{enumerate}
\item Let $A(t)$ be the age process corresponding to renewal process $N(t)$. Find the renewal equation for $E[A(t)]$ and compute $\lim_{t \to \infty}E[A(t)]$.
\item Consider successive flips of a fair coin. 
\begin{enumerate}
	\item Compute the mean no. of flips till the pattern HHTHHTT appears.
	\item Which pattern, HHTT or HTHT, requires a longer expected time to occur?
\end{enumerate}

\item Consider a delayed renewal process $\{N_D(t), t\geq 0\}$ whose first arrival has distribution $G$ and the others have distribution $F$. Let $m_D(t) = E[N_D(t)]$. Prove that
\[m_D(t) = G(t) + \int_0^t m(t-x)dF(x)\]
where $m(t) = \sum_{n=1}^\infty F^{*(n)}(t)$.

Let $A_D(t)$ be the age process of $N_D(t)$. Prove that if $F$ is non-lattice with $\int x^2 dF(x) < \infty$ and $\lim_{t \to \infty}t\overline{G}(t) \to 0$. Then
\[\lim_{t \to \infty}E[A_D(t)]  = \frac{\int_0^\infty x^2 dF(x)}{2\int_0^\infty x dF(x)}\]

\item Consider a GI/GI/1 queue. Interarrival times $\{A_n\}$ are iid and service times $\{S_n\}$ are iid with $E[S_n] < E[A_n] < \infty$. Let $V(t)$ be the virtual service time in the queue at time $t$ which is defined as the sum of remaining service time of all customers present in the system at time t. 
\begin{enumerate}
	\item Show that $V = \lim_{t \to \infty}\frac{\int_0^t V(u)du}{t}$ exists a.s and is constant.
	\item Let $D_n$ be the amount of time nth customer waits in the queue. Define
	\[W_Q = \frac{\sum_{k=1}^n D_k}{n}\]
	Show that $W_Q$ exists a.s. and is a constant.
	\item Let $E[A_n] = 1/\lambda$ and $Y$ have the same distribution as the service times $S_n$. Then show that
	\[V = \lambda E[Y]W_Q + \lambda E[Y^2]/2\] 
\end{enumerate}

\end{enumerate}
Mail any typos to \url{gautamshenoydmc@gmail.com}.
\end{document}