\documentclass[a4paper,english,12pt]{article}
\input{header}

%opening
\title{Stochastic Models for Multiserver management using load balancing}
\author{Team : Hemanth and Vinay Siram}

\begin{document}
\maketitle

\section{Motivation}
With increase in number of internet users, everyone is looking for faster and more efficient services. Load balancing addresses this problem by distributing the customers smartly over different available servers thus making the optimum use of resources and reducing the service time of the systems.   
In many practical applications involving multiple servers and queues such as banks where load balancing can be realised as distribution of work among all workers for faster service.\\

 \section{System Model}
Initially the arrival requests are treated as poisson process and the service times as exponential distributed and further generalised for renewal arrivals and general distributed service times.
\section{Problem Statement}
Instead of allocating the arrived request/customer to a server among a pool of servers randomly , we look for the servers with smaller queues for allocation. This not only decreases the service time but also also does optimum use of resources.  
\subsection{Justification}
The work done so far considers poisson arrivals and exponential distribution of the service times. We will try to consider models for renewal arrivals and generally 
distributed service time.\\
\section{Plan}
\begin{itemize}
	\item Week 1-2   :  Literature Survey  
	\item Week 3-5   :  Analysis for poisson arrivals and exponentally distributed service times.
	\item Week 6-7   :  Analysis for renewal arrivals and generally distributed service times.  
\end{itemize}


\end{document}