% !TEX spellcheck = en_U
% !TEX spellcheck = LaTeX
\documentclass[a4paper,11pt,english]{article}

\input{../header}

%\usepackage[]{geometry}
\newgeometry{left=1in,right=1in,top=1in,bottom=1in}

\title{Assignment 5: Martingales}
\author{SPQT - 2019}

\begin{document}
\maketitle
\textbf{Problem 1:} Let $\delta_1, \delta_2,\dots$ be independent with $\E[\delta_i] = 0$ Let $X_1 = \delta_1$ and $X_{n+1} = X_n + \delta_{n+1} f_n(X_1,X_2,\dots X_n)$. Suppose $X_n$ are integrable. Show that $\{X_n\}$ is a martingale. {\color{red} $f_n$ is not defined in the problem.}\\

\indent \textbf{Problem 2:} Let $\{X_n\}$ be a martingale with $\E[X_1] = 0$ and $\E[X_n^2] < \infty$.
\begin{enumerate}
\item Show that $\E[(X_{n+r} - X_n)^2] = \sum_{k=1}^r \E[(X_{n+k} - X_{n+k-1})^2]$.
\item Assume $\sum_n \E[(X_n - X_{n-1})^2] < \infty$. Prove that $X_n$ converges with probability 1.
\end{enumerate} 

\indent \textbf{Problem 3:} If $\{X_n\}$ is martingale bounded either above or below, then show that $\sup_n \E[|X_n|] < \infty$.\\

\indent \textbf{Problem 4:} Let $\{Y_n\}$ be i.i.d. with $\P\{Y_n = 1\} = p = 1-q = \P\{Y_n = -1\}$. Let $S_0 =0$, $S_n = Y_1+Y_2+\dots +Y_n$, $T = \inf\{S_n = -a$ or $S_n = b\}$. When $p \neq q$ show that 
\begin{align*}
\E[T] = \frac{b}{p-q} - \frac{a+b}{p-q} \frac{1-(p/q)^b}{1-(p/q)^{a+b}}.
\end{align*}

\indent \textbf{Problem 5:} Suppose $\{X_n\}$ is martingale. Let for some $\alpha$ $\E[|X_n|^\alpha] < \infty$ for all $n$. Show that 
\begin{align*}
\E \left[\max_{0\leq k \leq n} |X_n|\right] \leq \frac{\alpha}{1-\alpha} E[|X_n|^\alpha]^{\frac{1}{\alpha}}.
\end{align*}

\indent \textbf{Problem 6:} Show that a submartingale $\{X_n\}$ can be represented as $X_n = Y_n + Z_n$ where $\{Y_n\}$ is a martingale and $0 \leq Z_1 \leq Z_2 \leq \dots$. Hint: Take $X_0 = 0$, $\delta_n = X_n - X_{n-1}$ and $Z_n = \sum_{k=1}^n \E[\delta_k|\mathcal{F}_{k-1}]$. \\

\indent \textbf{Problem 7:}  Let $\{X_i\}$ be i.i.d. with $\P\{X_i = 0\} = \P\{X_i = 2\} = 1/2$. Check if $\{X_i\}$ is a martingale and if we can apply martingale stopping theorem. \\

\indent \textbf{Problem 8:} There are $n$ red balls, $n$ yellow balls and $m$ boxes. A red ball is kept in box $j$ with probability $p_j$ and a yellow ball with probability $q_j$ independently. Let $X$ be the number of boxes with one red and one yellow ball. Calculate $\E[X] = \mu $ and an exponential upper bound for $\E[|X-\mu|>b]$.
 \end{document}

